---
title: "nordpred"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nordpred}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Background

The Nordpred software package is a package of R and S-PLUS
functions for predicting trends in cancer incidence, using a special
version of the Age-Period-Cohort model (APC). The package is originally
written by Harald Fekjaer and Bjorn Moller at the Cancer Registry of Norway
as part of a larger cancer prediction project with financial support from
the Nordic Cancer Union (NCU).


# Case study

```{r}
library(nordpred)
```


## Data sets

Assume we have thre data sets, all with the following properties:

* All cells are numeric (the data set could be a matrix)
* Each row represents an age group
* Each column represents a time interval

In this example, all data concerns Norwegian males. Cancer incidence concerns colon cancer.

### 1 Observed population

The example data set `inpop1` gives observed population numbers for five year intervals from 1958 to 1997.
```{r}
knitr::kable(inpop1)
```

### 2 Predicted population

The example data set `inpop2` gives predicted population numbers for five year intervals from 1998 to 2022.
```{r}
knitr::kable(inpop2)
```

### 3 Cancer incidence

The example data set `indata` gives observed number of cancer incidences from the same time period as `inpop1`.
```{r}
knitr::kable(inpop2)
```

### Population

Functions in the nordpred package expect one data set for population data (both observed and predicted values).
```{r }
inpop <- cbind(inpop1, inpop2)
```

## Run predictions

Use `nordpred.estimate` to estimate a model 
```{r}
est <- nordpred.estimate(cases=indata, pyr = inpop, noperiod = 4, startestage = 5)
```

Then, use the outcome (a `nordpred` object) for prediction by `nordpred.prediction`
```{r}
res <- nordpred.prediction(est, startuseage = 6, cuttrend = c(0,.25,.5,.75,.75), recent=TRUE)
```
Or estimate and predict simultaneously using `nordpred` (same result)
```{r}
res <- nordpred(cases = indata, pyr = inpop, startestage = 5, startuseage = 6, 
                noperiods = 4, cuttrend = c(0, .25, .5, .75, .75))
```


## Choose number of periods

The "nordpred"-function can also choose number periods to base predictions on:
This is done by listing candidate number of periods in "noperiods". 
If the goodness of fit test is rejected based on the widest base, 
the first period is excluded etc.
```{r}
res <- nordpred(indata, inpop, startestage = 5, startuseage = 6, noperiods = 4:6,
                cuttrend=c(0, .25, .5, .75, .75))

```

## Poisson link

We can also use a poisson link function (instead of the powerlink as used in the nordpred predictions).
```{r}
est2 <- nordpred.estimate(indata, inpop, 4, 5, linkfunc = "poisson")
res2 <- nordpred.prediction(est2, startuseage = 6, cuttrend = c(0, .25, .5, .75, .75), recent = TRUE)
```

## Results
```{r}
print(res)
nordpred.getpred(res)
summary(res, printpred = FALSE)
```


## Get results with standardisation:
```{r}
## World population standard
wstand <- c(0.12, 0.1, 0.09, 0.09, 0.08, 0.08, 0.06, 0.06, 0.06, 0.06,0.05, 
            0.04, 0.04, 0.03, 0.02, 0.01, 0.005, 0.005)
            
round(nordpred.getpred(res, incidence = TRUE, standpop = NULL), 2)
round(nordpred.getpred(res, incidence = TRUE, standpop = wstand), 2)

```

## Plot results:
```{r}
plot(res, standpop = wstand)
```


## Plot results with power5 and poisson links:
```{r}
plot(res2, standpop = wstand)
plot(res, new = FALSE, lty = c(1, 2), standpop = wstand)
```

# Different cut trend scenarios
using average drift (recent = FALSE)
```{r}
plot(nordpred.prediction(est, startuseage = 6, cuttrend = c(0, 0, 0, 0, 0), 
    recent = FALSE), standpop = wstand, new = TRUE)
plot(nordpred.prediction(est, startuseage = 6, cuttrend = c(1, 1, 1, 1, 1), 
    recent = FALSE), standpop = wstand, new = FALSE, lty = c(1, 2))
plot(nordpred.prediction(est, startuseage = 6, cuttrend = c(0, .25, .5, .75, .75), 
    recent = FALSE), standpop = wstand, new = FALSE, lty = c(1, 4))
```


